---
title: "Getting started with phytaxr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with phytaxr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`phytaxr` provides a reproducible pipeline for cleaning and resolving
phytoplankton taxon names against [WoRMS](https://www.marinespecies.org).
This vignette walks through the full workflow, from a raw list of names to a
resolved, taxonomy-enriched data frame.

## Installation

```{r install}
# From the package source directory:
devtools::install()
```

## Workflow overview

The pipeline has three stages:

1. **Clean** — normalise raw names step by step (encoding, abbreviations,
   uncertain taxa, separator expansion, generic names)
2. **Resolve** — match cleaned names against WoRMS and GBIF (exact, taxamatch,
   fuzzy minor correction) and retrieve full classification
3. **Review** — interactively resolve any remaining unmatched taxa

## 1. Load the package

```{r load}
library(phytaxr)
library(dplyr)
```

## 2. Prepare your input

Start from a character vector of raw taxon names as they appear in your
original data — abbreviations, common names, and OCR artefacts included.

```{r input}
raw_taxa <- c(
  "Thalassiosira weissflogii",
  "Chaetoceros sp.",
  "diatom",
  "Skeletonema Costatum",        # wrong capitalisation
  "Nitzchia longissima",         # typo: Nitzchia -> Nitzschia
  "Gymnodinium / Gyrodinium",    # slash entry
  "Coscinodiscus cf. radiatus"   # uncertain identification
)
```

## 3. Clean the names step by step

The cleaning pipeline is composed of individual functions that you can call
sequentially. Each function takes a data frame with at minimum a `taxon_clean`
column and returns it modified. Use `tibble::tibble()` to initialise the
input, then pipe through each step:

```{r clean}
library(tibble)

df <- tibble(
  taxon_original = raw_taxa,
  taxon_clean    = raw_taxa
)

df <- df |>
  normalize_characters()           |>  # 1. encoding, diacritics, whitespace
  process_taxonomic_prefixes()     |>  # 2. cf., aff., s.l., s.str., etc.
  process_incertae_entries()       |>  # 3. incertae sedis
  process_sp_entries()             |>  # 4. sp., spp., sp. nov.
  process_bracket_entries()        |>  # 5. parenthesised qualifiers
  move_size_to_epithet()           |>  # 6. size annotations
  process_epithet_entries()        |>  # 7. epithets
  normalize_infraspecific_ranks()  |>  # 8. var., subsp., f., etc.
  remove_dots()                    |>  # 9. stray dots
  move_reproductive_structures()   |>  # 10a. cysts, spores, …
  move_uncertainty_descriptors()   |>  # 10b. unidentified, unknown, …
  move_morphological_descriptors() |>  # 10c. centric, pennate, …
  move_with_descriptors()          |>  # 10d. "with …" phrases
  move_formia_to_epithet()         |>  # 10e. forma
  move_commas_to_epithet()         |>  # 10f. post-comma strings
  move_authors_to_epithet()        |>  # 10g. authorships
  remove_sp_tokens()               |>  # 11. sp/spp/ssp tokens in epithet
  split_separator_entries()        |>  # 12. unfold /, &, + entries
  process_generic_taxa()           |>  # 13. vernacular → taxon name
  clean_trailing_hyphens()             # 14. stray hyphens
```

## 4. Resolve against WoRMS and GBIF

`resolve_taxonomy()` is the main entry point for the external resolution
stage. It runs the full sequential pipeline on a data frame that already
has a `taxon_clean` column (as produced by the cleaning step above, or
directly on raw names):

```{r resolve}
result <- resolve_taxonomy(raw_taxa)
```

The output is a tibble with one row per input name (slash entries are
expanded) and columns for matched and accepted names, AphiaIDs, taxonomic
status, resolution method, and the full classification from kingdom to forma.

```{r inspect}
# Resolution summary
result |>
  count(resolution_method, sort = TRUE)

# Unresolved entries
result |>
  filter(is.na(matched_aphiaid)) |>
  select(taxon_original, taxon_clean)
```

## 5. Interactive fuzzy review

For taxa that could not be resolved automatically, `process_fuzzy_batch()`
presents cascading WoRMS suggestions and collects your choices interactively.
Progress is saved to a checkpoint file after each accepted entry so sessions
can be interrupted and resumed.

```{r fuzzy}
# Build vocabularies from already-resolved taxa to power edit-distance search
genus_vocab   <- build_genus_vocabulary(result)
epithet_vocab <- build_epithet_vocabulary(result)

# Run the interactive loop (only processes rows where matched_aphiaid is NA)
result <- process_fuzzy_batch(
  df              = result,
  genus_vocab     = genus_vocab,
  epithet_vocab   = epithet_vocab,
  checkpoint_file = "phytaxr_checkpoint.rds"
)
```

Each unresolved entry shows ranked candidates grouped by confidence tier
(good, longshot, genus fallback). You can:

- Enter a number to accept a suggestion
- `m` to enter an AphiaID or name manually
- `e` to flag for expert review
- `f` to flag for removal
- `s` to skip
- `c` to stop the session

## 6. Save results

```{r save}
readr::write_csv(result, "taxonomy_resolved.csv")
```

## Vernacular name dictionary

`phytaxr` ships with an internal dictionary of common phytoplankton vernacular
names (e.g. `"diatom"` → `"Bacillariophyceae"`). The dictionary is stored in
`R/sysdata.rda` and loaded automatically. Session-level changes are kept in
memory and lost on restart; to make them permanent, edit
`data-raw/vernacular.R` and re-run it.

```{r vernacular}
# List all entries
vernacular_list()

# Look up a single name (case-insensitive)
vernacular_lookup("diatom")
#> [1] "Bacillariophyceae"

# Search by pattern (in keys, values, or both)
vernacular_search("diatom")

# Add a new entry (session only)
vernacular_add("green alga", "Chlorophyta")

# Update an existing entry (session only)
vernacular_update("diatom", "Bacillariophyta")

# Remove an entry (session only)
vernacular_remove("green alga")
```
