---
title: "Getting started with phytaxr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with phytaxr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`phytaxr` provides a reproducible workflow for cleaning, standardising, and
resolving phytoplankton taxon names against external taxonomic databases,
primarily WoRMS and GBIF.

This vignette reflects the current package design. In particular, Stage 1
(cleaning) is built from **atomic** functions: each cleaning function can be
called independently on a data frame that contains only a `taxon` column.
Internally, `ensure_cleaning_schema()` creates the working columns required by
the cleaning pipeline when they are missing.

## Installation

```{r install}
# install.packages("remotes")
remotes::install_github("RayonVina/phytaxr")
```

## Workflow overview

The package is organised into three stages:

1. **Cleaning** — normalise raw names and extract qualifiers into a structured
   `tax_epithet` column.
2. **Automatic resolution** — match cleaned names against WoRMS and GBIF and
   retrieve accepted names, AphiaIDs, taxonomic status, and taxonomy.
3. **Semi-automatic fuzzy review** — interactively review unresolved names in
   batches using ranked suggestions.

## Load the package

```{r load}
library(phytaxr)
library(dplyr)
library(tibble)
```

## Prepare your input

Start with a data frame containing a raw taxon column named `taxon`.

```{r input}
df <- tibble(
  taxon = c(
    "Chaetoceros debilis",
    "Thalassiosira weissflogii",
    "C. diatom sp.",
    "Gymnodinium cf. catenatum",
    "Pseudo nitzschia delicatissima",
    "Skeletonema costatum [chain-forming]",
    "Gymnodinium / Gyrodinium"
  )
)
```

## Stage 1: cleaning

### Recommended workflow

For most use cases, the recommended entry point is the Stage 1 wrapper:

```{r clean-wrapper}
df_clean <- run_cleaning_pipeline(df)
```

This applies the full cleaning workflow in the package's recommended sequence
and returns a data frame with the standard cleaning output columns, including:

- `taxon` — original raw input
- `taxon_clean` — cleaned scientific name
- `tax_epithet` — structured qualifiers extracted during cleaning
- `uncertain` — logical flag for names that could not be fully normalised

### Atomic functions

A key feature of the current implementation is that cleaning functions are
**atomic**. Each function can be used independently, even if no previous
cleaning step has been run.

For example, the following calls work on their own:

```{r atomic-example}
process_incertae_entries(
  tibble(taxon = "Chaetoceros cf. debilis")
)

process_sp_entries(
  tibble(taxon = "Chaetoceros sp.")
)

move_size_to_epithet(
  tibble(taxon = "Coscinodiscus 20-30 µm")
)
```

This is possible because `ensure_cleaning_schema()` checks whether the required
working columns are present and creates them when needed. Most users will not
need to call `ensure_cleaning_schema()` directly, but it is a central part of
how Stage 1 now works internally.

### Modular use

You can still call the cleaning functions one by one when you want to inspect
intermediate results or customise the workflow:

```{r clean-modular}
df_clean <- df |>
  normalize_characters() |>
  process_taxonomic_prefixes() |>
  process_incertae_entries() |>
  process_sp_entries() |>
  process_bracket_entries() |>
  move_size_to_epithet() |>
  process_epithet_entries() |>
  normalize_infraspecific_ranks() |>
  remove_dots() |>
  move_reproductive_structures() |>
  move_uncertainty_descriptors() |>
  move_morphological_descriptors() |>
  move_with_descriptors() |>
  move_formia_to_epithet() |>
  move_commas_to_epithet() |>
  move_authors_to_epithet() |>
  remove_sp_tokens() |>
  split_separator_entries() |>
  process_generic_taxa() |>
  clean_trailing_hyphens() |>
  remove_short_interstitial_tokens() |>
  apply_vernacular_corrections()
```

These functions no longer depend on a strict execution order just to work
correctly in isolation. However, when you want a full, reproducible cleaning
workflow, `run_cleaning_pipeline()` remains the recommended approach because it
applies the intended sequence of transformations.

## Stage 2: automatic resolution

Once names have been cleaned, run the automatic resolution pipeline:

```{r resolve-wrapper}
df_res <- run_resolution_pipeline(df_clean)
```

This wrapper applies the current sequence of automatic resolution steps:

1. `search_worms_priority()`
2. `search_worms_taxamatch()`
3. `search_gbif_strict()`
4. `resolve_taxonomic_status()`
5. `search_worms_fuzzy_minor()`
6. `get_taxonomy()`

The output adds resolution metadata such as:

- `matched_name`
- `matched_aphiaid`
- `accepted_name`
- `accepted_aphiaid`
- `taxonomic_status`
- `resolution_method`
- `resolution_notes`
- taxonomic ranks from `kingdom` to `forma`

You can inspect the result, for example:

```{r inspect}
df_res |>
  count(resolution_method, sort = TRUE)

df_res |>
  filter(is.na(matched_aphiaid)) |>
  select(taxon, taxon_clean)
```

Resolution functions can also be called individually when needed:

```{r resolve-modular}
df_res <- search_worms_priority(df_clean)
df_res <- search_worms_taxamatch(df_res)
df_res <- search_gbif_strict(df_res)
df_res <- resolve_taxonomic_status(df_res)
df_res <- search_worms_fuzzy_minor(df_res)
df_res <- get_taxonomy(df_res)
```

## Stage 3: semi-automatic fuzzy review

Names that remain unresolved after Stage 2 can be reviewed interactively.

First, build vocabularies from the names that were already resolved:

```{r fuzzy-vocab}
genus_vocab <- build_genus_vocabulary(df_res)
epithet_vocab <- build_epithet_vocabulary(df_res)
```

Then run the interactive batch processor:

```{r fuzzy}
df_final <- process_fuzzy_batch(
  df = df_res,
  genus_vocab = genus_vocab,
  epithet_vocab = epithet_vocab,
  taxon_col = "taxon",
  taxon_clean_col = "taxon_clean",
  batch_size = 10,
  checkpoint_file = "phytaxr_step3_checkpoint.rds",
  min_similarity = 0.85,
  max_suggestions = 15,
  edit_max_dist = 3,
  timeout_sec = 15
)
```

The function processes only unresolved rows and saves progress to a checkpoint
file so that interrupted sessions can be resumed later.

## Vernacular dictionary

`phytaxr` includes an internal dictionary of common phytoplankton vernacular
names (for example, `"diatom"` → `"Bacillariophyceae"`).

Two cleaning functions use this dictionary at different points in the workflow:

- `process_generic_taxa()` performs an exact-match lookup on the full
  `taxon_clean` string.
- `apply_vernacular_corrections()` performs a prefix-based correction and
  preserves any trailing epithet.

Examples:

```{r vernacular}
vernacular_list()

vernacular_lookup("diatom")

vernacular_search("diatom")

vernacular_add("green alga", "Chlorophyta")

vernacular_update("diatom", "Bacillariophyta")

vernacular_remove("green alga")
```

Session-level changes are temporary. To make dictionary edits permanent, update
`data-raw/vernacular.R` and re-run that script.

## Save the results

```{r save}
readr::write_csv(df_final, "taxonomy_resolved.csv")
```