---
title: "Getting started with phytaxr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with phytaxr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`phytaxr` provides a reproducible pipeline for cleaning and resolving
phytoplankton taxon names against [WoRMS](https://www.marinespecies.org).
This vignette walks through the full workflow, from a raw list of names to a
resolved, taxonomy-enriched data frame.

## Installation

```{r install}
# From the package source directory:
devtools::install()
```

## Workflow overview

The pipeline has three stages:

1. **Clean** — normalise raw names step by step (encoding, abbreviations,
   uncertain taxa, separator expansion, generic names)
2. **Resolve** — match cleaned names against WoRMS and GBIF (exact, taxamatch,
   fuzzy minor correction) and retrieve full classification
3. **Review** — interactively resolve any remaining unmatched taxa

## 1. Load the package

```{r load}
library(phytaxr)
library(dplyr)
```

## 2. Prepare your input

Start from a character vector of raw taxon names as they appear in your
original data — abbreviations, common names, and OCR artefacts included.

```{r input}
raw_taxa <- c(
  "Thalassiosira weissflogii",
  "Chaetoceros sp.",
  "diatom",
  "Skeletonema Costatum",        # wrong capitalisation
  "Nitzchia longissima",         # typo: Nitzchia -> Nitzschia
  "Gymnodinium / Gyrodinium",    # slash entry
  "Coscinodiscus cf. radiatus"   # uncertain identification
)
```

## 3. Clean the names step by step

The cleaning pipeline is composed of individual functions that you can call
sequentially. Each function takes a data frame with at minimum a `taxon_clean`
column and returns it modified. Use `tibble::tibble()` to initialise the
input, then pipe through each step:

```{r clean}
library(tibble)

df <- tibble(
  taxon_original = raw_taxa,
  taxon_clean    = raw_taxa
)

df <- df |>
  normalize_characters()           |>  # 1. encoding, diacritics, whitespace
  process_taxonomic_prefixes()     |>  # 2. cf., aff., s.l., s.str., etc.
  process_incertae_entries()       |>  # 3. incertae sedis
  process_sp_entries()             |>  # 4. sp., spp., sp. nov.
  process_bracket_entries()        |>  # 5. parenthesised qualifiers
  move_size_to_epithet()           |>  # 6. size annotations
  process_epithet_entries()        |>  # 7. epithets
  normalize_infraspecific_ranks()  |>  # 8. var., subsp., f., etc.
  remove_dots()                    |>  # 9. stray dots
  move_reproductive_structures()   |>  # 10a. cysts, spores, …
  move_uncertainty_descriptors()   |>  # 10b. unidentified, unknown, …
  move_morphological_descriptors() |>  # 10c. centric, pennate, …
  move_with_descriptors()          |>  # 10d. "with …" phrases
  move_formia_to_epithet()         |>  # 10e. forma
  move_commas_to_epithet()         |>  # 10f. post-comma strings
  move_authors_to_epithet()        |>  # 10g. authorships
  remove_sp_tokens()               |>  # 11. sp/spp/ssp tokens in epithet
  split_separator_entries()        |>  # 12. unfold /, &, + entries
    process_generic_taxa()           |>  # 13. vernacular → taxon name (exact match)
  clean_trailing_hyphens()         |>  # 14. stray hyphens
  remove_short_interstitial_tokens() |> # 15. 1–2 char genus abbreviation artefacts
  apply_vernacular_corrections()        # 16. prefix-based name correction from dictionary
```

## 4. Resolve against WoRMS and GBIF

`run_resolution_pipeline()` runs the full sequential resolution pipeline on a
data frame that already has a `taxon_clean` column:

```{r resolve}
df_res <- run_resolution_pipeline(df)
```

The output adds columns for matched and accepted names, AphiaIDs, taxonomic
status, resolution method, and the full classification from kingdom to forma.

```{r inspect}
# Resolution summary
df_res |>
  dplyr::count(resolution_method, sort = TRUE)

# Unresolved entries
df_res |>
  dplyr::filter(is.na(matched_aphiaid)) |>
  dplyr::select(taxon_clean)
```

## 5. Interactive fuzzy review

For taxa that could not be resolved automatically, `process_fuzzy_batch()`
presents cascading WoRMS suggestions and collects your choices interactively.
Progress is saved to a checkpoint file after each accepted entry so sessions
can be interrupted and resumed.

```{r fuzzy}
# Build vocabularies from already-resolved taxa to power edit-distance search
genus_vocab   <- build_genus_vocabulary(result)
epithet_vocab <- build_epithet_vocabulary(result)

# Run the interactive loop (only processes rows where matched_aphiaid is NA)
result <- process_fuzzy_batch(
  df              = result,
  genus_vocab     = genus_vocab,
  epithet_vocab   = epithet_vocab,
  checkpoint_file = "phytaxr_checkpoint.rds"
)
```

Each unresolved entry shows ranked candidates grouped by confidence tier
(good, longshot, genus fallback). You can:

- Enter a number to accept a suggestion
- `m` to enter an AphiaID or name manually
- `e` to flag for expert review
- `f` to flag for removal
- `s` to skip
- `c` to stop the session

## 6. Save results

```{r save}
readr::write_csv(result, "taxonomy_resolved.csv")
```

## Vernacular name dictionary

`phytaxr` ships with an internal dictionary of common phytoplankton vernacular
names (e.g. `"diatom"` → `"Bacillariophyceae"`). The dictionary is stored in
`R/sysdata.rda` and loaded automatically. Session-level changes are kept in
memory and lost on restart; to make them permanent, edit
`data-raw/vernacular.R` and re-run it.
`process_generic_taxa()` (step 13) performs an exact-match lookup of the full
`taxon_clean` string against the dictionary. `apply_vernacular_corrections()`
(step 16) complements it with a prefix-based search: if a dictionary key is
found at the *start* of `taxon_clean`, it replaces it with the canonical form
while preserving any trailing epithet (e.g. `"pseudo nitzschia delicatissima"`
→ `"Pseudo-nitzschia delicatissima"`).

```{r vernacular}
# List all entries
vernacular_list()

# Look up a single name (case-insensitive)
vernacular_lookup("diatom")
#> [1] "Bacillariophyceae"

# Search by pattern (in keys, values, or both)
vernacular_search("diatom")

# Add a new entry (session only)
vernacular_add("green alga", "Chlorophyta")

# Update an existing entry (session only)
vernacular_update("diatom", "Bacillariophyta")

# Remove an entry (session only)
vernacular_remove("green alga")
```
